---
title: "Evaluating Serve Tactics based on ATP Performance in 2022"
author: "Jarrett Markman"
date: '2023-02-23'
output: html_document
---

#### Abstract

ATP players like Maxime Cressy and Alexander Zverev have made waves in the past few years, hitting only first serves, including first serves on their second serves. Is it a viable strategy for players? Or is the normal way of play hitting first and second serves still dominant? Using expected mean I can calculate the win probability before a single point for hitting first serves, normal serves, or second serves.

Each serve and each point after hitting the serve in are mutually exclusive events. So on a given point there are 4 mutually exclusive events - hitting the first serve in, winning the point on a first serve in, hitting the second serve in after missing the first serve, and winning the point on a second serve in. 

It is also important to note that the probability of 2 mutually exclusive events both occurring (p(A) and p(B)) is equivalent to p(A) x p(B), and the probability of one of 2 mutually exclusive events occurring (p(A) or p(B)) is p(A) + p(B). In this case the events can be split up in 3 ways. winning the point hitting only first serves, winning the point hitting normal serves, and winning the point hitting only second serves.

These 3 events translate to:

The likelihood of winning one point hitting only first serves = 
p(win the point on the first serve) = (first serve in %) and (first serve win %) OR
p(win the point on the second serve) = (first serve out % and first serve in %) and (first serve win %). 

Which mathematically translates to: 

p(point WP) = p(first serve win) + p(second serve win) = ((first serve in %) x (first serve win %)) + ((1 - first serve in %) x (first serve in %) x (first serve win %))

The likelihood of winning one point hitting normal serves = 
p(win the point on the first serve) = (first serve in %) and (first serve win %) OR
p(win the point on the second serve) = (first serve out % and second serve in %) and (second serve win %)

Which mathematically translates to:

p(point WP) = p(first serve win) + p(second serve win) = ((first serve in %) x (first serve win %)) + ((1 - first serve in %)  x (second serve in %) x (second serve win %))

The likelihood of winning one point hitting only second serves = 
p(win the point on the first serve) = (second serve in %) and (second serve win %) OR
p(win the point on the second serve) = (second serve out % and second serve in %) and (second serve win %)

Which mathematically translates to:

p(point WP) = p(first serve win) + p(second serve win) = ((second serve in %) x (second serve win %)) + ((1 - second serve in %) and (second serve in %) x (second serve win %))

Lets say a given player hits their first serve in 60% of the time, wins 90% of first serve points, hits their second serve in 90% of the time, and wins 50% of the points.

first serve in % = 60% = 0.6

first serve win % = 90% = 0.9

second serve in % = 90% = 0.9

second serve win % = 50% = 0.5

What is their point win probability hitting only first serves?

WP = (.6) x (.9) + (1 - .6) x (.6) x (.9) = .54 + .216 = .756,

meaning that over the course of 1000 points, hitting only first serves would win 756 points.

What is their point win probability hitting normal serves?

WP = (.6) x (.9) + (1 - .6) x (.9) x (.5) = .54 + .18 = .720,

meaning that over the course of 1000 points, hitting normal serves would win 720 points.

What is their point win probability hitting second serves?

WP = (.9) x (.5) + (1 - .9) x (.9) x (.5) = .45 + .045 = .495,

meaning that over the course of 1000 points, hitting only second serves would win 495 points. 

Mathematically and logically it would make sense for this player to hit only first serves. This is due in part to the fact that they hit their first serve in more often than they win second serve points. They are 1.2 times more likely to hit their first serve in with a 90% win rate than win their second serve point, hitting it in 90% of the time, meaning this player can expect to have won 12 points hitting only first serves, and only 10 points hitting a first and second serve. 

For measuring each players serve ability, I will be using expected mean to calculate their point win probability under 3 different circumstances: the likelihood they win a point with the intention of hitting two first serves, the likelihood they win a point with the intention of hitting a first and second serve, and the likelihood they win a point with the intention of hitting two second serves.

```{r setup, echo=F, message=F}
options(dplyr.summarise.inform = FALSE)
```

``` {r, warning = FALSE, messages = FALSE, echo = FALSE, include = FALSE}
#load libraries
library("tidyverse")
library("readr")
library("gt")
```

``` {r, echo = FALSE, message = FALSE}
#weighted probability for first serves
data <- read_csv("match data/atp_matches_2022.csv")
```

``` {r, echo = FALSE, message = FALSE}
#aggregate winners data
winners_pulled <- data %>%
  select(winner_name, w_svpt, w_1stIn, w_1stWon, w_2ndWon, w_df, w_ace)
winners_renamed <- winners_pulled %>%
  rename(
    player = winner_name,
    first_points = w_svpt,
    first_in = w_1stIn,
    first_won = w_1stWon,
    second_won = w_2ndWon,
    dfs = w_df,
    ace = w_ace
  )
winners <- winners_renamed %>%
  mutate(
    second_points = first_points - first_in,
    second_in = second_points - dfs
  )
```

``` {r, echo = FALSE, message = FALSE}
#aggregate losers data
losers_pulled <- data %>%
  select(loser_name, l_svpt, l_1stIn, l_1stWon, l_2ndWon, l_df, l_ace)
losers_renamed <- losers_pulled %>%
  rename(
    player = loser_name,
    first_points = l_svpt,
    first_in = l_1stIn,
    first_won = l_1stWon,
    second_won = l_2ndWon,
    dfs = l_df,
    ace = l_ace
  )
losers <- losers_renamed %>%
  mutate(
    second_points = first_points - first_in,
    second_in = second_points - dfs
  )
```

``` {r, echo = FALSE, message = FALSE}
#combine data
combined <- bind_rows(winners, losers) %>%
  filter(!is.na(first_points)) %>%
  group_by(player) %>%
  summarise(
    first_points = sum(first_points),
    first_points_in = sum(first_in),
    first_points_won = sum(first_won),
    second_points = sum(second_points),
    second_points_in = sum(second_in),
    second_points_won = sum(second_won),
    total_dfs = sum(dfs),
    total_aces = sum(ace)
  ) %>%
  filter(first_points > 100)
```

``` {r, echo = FALSE, message = FALSE}
#player data frame with serve point averages
players <- combined %>%
  summarise(
    player = player,
    first_serves_in_pct = first_points_in/first_points,
    first_serves_won_pct = first_points_won/first_points_in,
    second_serves_in_pct = second_points_in/second_points,
    second_serves_won_pct = second_points_won/second_points_in
  )
```

``` {r, echo = FALSE, message = FALSE}
#data frame with player probabilities for points won
player_probabilities <- players %>%
  summarise(
    player = player,
    likelihood_first_won = first_serves_in_pct * first_serves_won_pct,
    likelihood_second_won = second_serves_in_pct * second_serves_won_pct,
    likelihood_first_missed_second_first_won = (1-first_serves_in_pct) * 
      likelihood_first_won,
    likelihood_first_missed_second_second_won = (1-first_serves_in_pct) * 
      likelihood_second_won,
    likelihood_first_second_missed_second_second_won = (1-second_serves_in_pct) * 
      likelihood_second_won,
    first_weighted = likelihood_first_won + likelihood_first_missed_second_first_won,
    normal_weighted = likelihood_first_won + likelihood_first_missed_second_second_won,
    second_weighted = likelihood_second_won + likelihood_first_second_missed_second_second_won
  )
#x_weighted - meaning the point win probability based on x
```

``` {r, echo = FALSE, message = FALSE}
#players who should hit only first serves
first_servers <- player_probabilities %>%
  filter(first_weighted > normal_weighted & first_weighted > second_weighted) %>%
  summarise(Player = player, 
         "Point Win Probability Hitting First Serves" = paste(round(first_weighted, 3) * 100, "%"), 
         "Point Win Probability Hitting Normal Serves" = paste(round(normal_weighted, 3) * 100, "%"), 
         "Point Win Probability Hitting Second Serves" = paste(round(second_weighted, 3) * 100, "%"))
```

``` {r, echo = FALSE, message = FALSE}
#players who should hit normal serves
normal_servers <- player_probabilities %>%
  filter(normal_weighted > first_weighted & normal_weighted > second_weighted) %>%
  summarise(Player = player, 
            "Point Win Probability Hitting First Serves" = paste(round(first_weighted, 3) * 100, "%"), 
            "Point Win Probability Hitting Normal Serves" = paste(round(normal_weighted, 3) * 100, "%"), 
            "Point Win Probability Hitting Second Serves" = paste(round(second_weighted, 3) * 100, "%"))
```

``` {r, echo = FALSE, message = FALSE}
#players who should hit only second serves
second_servers <- player_probabilities %>%
  filter(second_weighted > first_weighted & second_weighted > normal_weighted) %>%
  summarise(Player = player, 
            "Point Win Probability Hitting First Serves" = paste(round(first_weighted, 3) * 100, "%"), 
            "Point Win Probability Hitting Normal Serves" = paste(round(normal_weighted, 3) * 100, "%"), 
            "Point Win Probability Hitting Second Serves" = paste(round(second_weighted, 3) * 100, "%"))
```


``` {r, echo = FALSE}
#first servers
first_servers %>%
  gt() %>%
  tab_spanner(
    label = "Players who should hit only First Serves",
    columns = c(Player, `Point Win Probability Hitting First Serves`,
                `Point Win Probability Hitting Normal Serves`,
                `Point Win Probability Hitting Second Serves`)
  ) %>%
  tab_options(table.background.color = "lightblue2") %>%
  cols_align(
    align = c("center"), columns = everything()) %>%
  opt_table_font(font = "Roboto Condensed") %>%
  opt_table_outline() %>%
  tab_source_note(
    source_note = "Jarrett Markman | Data: Jeff Sackmann - GitHub"
  )
```

It's no surprise that someone like Alexander Zverev is on this list, given how much he struggles with his second serve. The reason why many of the players on this list are here is due to the fact that their second serves are very weak. For most these players there is larger than 10% separated between the win probability hitting first serves than the win probability hitting second serves. This can be due to a multitude of factors, such as an extremely low second serve win %, or an extremely low second serve in %, as well as the server possibly having such great first serve numbers. Zverev is a great example of this, because his first serve is great, however, his second serve is extremely inconsistent, often resulting in double faults, or opponents pouncing on the poor serve. 

``` {r, echo = FALSE}
#second servers
second_servers %>%
  gt() %>%
  tab_spanner(
    label = "Players who should hit only Second Serves",
    columns = c(Player, `Point Win Probability Hitting Second Serves`,
                `Point Win Probability Hitting Normal Serves`,
                `Point Win Probability Hitting First Serves`)
  ) %>%
  tab_options(table.background.color = "lightblue2") %>%
  cols_align(
    align = c("center"), columns = everything()) %>%
  opt_table_font(font = "Roboto Condensed") %>%
  opt_table_outline() %>%
  tab_source_note(
    source_note = "Jarrett Markman | Data: Jeff Sackmann - GitHub"
  )
```

This is most likely due in part for the opposite reason as only first servers. It's possible that they have a first serve that struggles to go in, or one that is ineffective, especially in comparison to a second serve that is much more likely to go in. It's possible that some of these players have very good second serves or very poor first serves, or both. 

``` {r, echo = FALSE}
#normal servers
normal_servers %>%
  gt() %>%
  tab_spanner(
    label = "Players who should hit only Normal Serves",
    columns = c(Player, `Point Win Probability Hitting Normal Serves`,
                `Point Win Probability Hitting First Serves`,
                `Point Win Probability Hitting Second Serves`)
  ) %>%
  tab_options(table.background.color = "lightblue2") %>%
  cols_align(
    align = c("center"), columns = everything()) %>%
  opt_table_font(font = "Roboto Condensed") %>%
  opt_table_outline() %>%
  tab_source_note(
    source_note = "Jarrett Markman | Data: Jeff Sackmann - GitHub"
  )
```

Most players based on last year's data would appear to find no additional success by changing up their service strategy. This is most likely due in part to having a good/solid first serve, and a good/solid second serve, and not having an outlying statistic causing their percentages to drop - for example a very low first serve in % (ie. 40%) would most likely indicate that they player should not hit many first serves, because they don't go in often. 

#### Conclusion

For some players, sacrificing the increased likelihood of a double fault is worth it to make up for the low chance of winning a second serve point - whether there is a poor second serve in % or second serve win %. For others, forgoing an inconsistent/ineffective first serve to prevent any double faults and being consistent with each serve is worth it to make up for that inconsistent/ineffective first serve. However, for many players - barring an extremely good or poor first or second serve - it makes the most sense for them to continue to hit normal first and second serves, because it gives them the best chance to win points on a regular basis.